Extracting data from vector figures in scholarly articles

### 2017-08-27 submission/manuscript.Rmd
Update manuscript.Rmd

@@

@@

Based on the plot body, absolute locations of the individual data points are extracted. Not all vector images are created in a similar way, but in the simplest scenario with data points depicted as circles, the vector gives three parameters: the $x$ coordinate of the centre, the $y$ coordinate of the centre, and the radius $r$. As such, for a simple circle the underlying vector code might look as follows:
```html
<circle cx="103.71" cy="121.22" r="25.234" fill-opacity="0" stroke="#cf1d35" stroke-width=".26458"/>
```
This information can be readily extracted after isolating the vector figure from a PDF file. The current alpha software is primarily developed to operate on circles of similar size within one plot but can be extended for data depicted in other ways.

In order to make the absolute locations of the shapes represent the original data points as accurately as possible, they are mapped onto the identified x- and y-axis. Although absolute locations retain the relative relations between the individual data points, they are not representative of the original data. As such, the software extracts the locations of the axes and their respective scales. Subsequently, the location of the x-axis and y-axis, combined with the information about the scale, are combined to remap the absolute locations on the canvas into the original data points. The current alpha software assumes a linear scale, but logarithmic scales could be incorporated at a future stage.
<!-- verify with PMR -->

@@

In order to extract data from vector figures with the software `norma`, `r stepsnr` steps are taken. First, the user needs to organize all original PDFs into one folder. Second, this folder needs to be converted to a `cproject` structure. The `cproject` structure normalizes the contents for each paper into a `ctree`, such that subsequent operations are trivial to standardize (and extensions can be applied relatively easily). For example, the root folder might contain `ctree1.pdf`, but after transforming the root folder into a `cproject` it contains a folder `ctree1/` with `fulltext.pdf`. By running the command 
```bash
java -jar bin/norma-0.5.0-SNAPSHOT-jar-with-dependencies.jar --project corpus-raw  --fileFilter '.*/(.*).pdf' --makeProject '(\1)/fulltext.pdf'
```
the folder `corpus-raw` (`--project corpus-raw`) is restructured into a `cproject` structure, containing a folder for each PDF file (`--fileFilter '.*/(.*).pdf' --makeProject '(\1)/fulltext.pdf'`). This results in the following folder structure:
```bash
cproject/
├── ctree1
│   └── fulltext.pdf
├── ctree2
│   └── fulltext.pdf
...
└─── ctreeN
    └── fulltext.pdf
```

@@

As the results indicate, vector figures are a fruitful and feasible resource for data extraction. Based on initial alpha software of `norma` to extract data from vector figures, we correctly extracted data from `r round((dim(vec_dat[vec_dat$x_ax_mapped == 'yes' & vec_dat$y_ax_mapped == 'yes' & vec_dat$datapoints_extracted == vec_dat$datapoints_manual, ])[1] / dim(vec_dat)[1]) * 100, 0)`% of funnel plots for which data was extracted. This is a very strict assessment, considering our manual investigation depicted that four extracted datasets only required a simple reversal of one axis to be 100% accurate (i.e., 1.23 should be -1.23, etc.), adding a constant to all data points on an axis to adjust an incorrect mapping, or by rescaling an axis to fit to the logarithmic scale of one axis. If these manual corrections for four funnel plots are made, `r round(((dim(vec_dat[vec_dat$x_ax_mapped == 'yes' & vec_dat$y_ax_mapped == 'yes' & vec_dat$datapoints_extracted == vec_dat$datapoints_manual, ])[1] + 4) / dim(vec_dat)[1]) * 100, 0)`% of all funnel plots for which data were correctly extracted.

>>>>

PMR It is common for authors to communicate their results in graphical diagrams, because these can transmit message-rich views of the data. Common examples are x-y plots of data points for the dependence of one variable (y) on another (x) or two covarying quantities (scatterplots). Authors will use a plotting program to obtain meaningful views of the variance and then submit the diagram as an x-y plot. What is generally not realised it that it may be possible to reconstruct the original x-y data from the plot.

@@

PMR. Conventionally the plot is designed for inclusion in a PDF document with high resolution but also display on a raster (pixel) screen with restricted to an area of (say) 500 pixels (x) by 700 pixels (y). To extract data points manually an author may have to measure the coordinates either on printed pages using a ruler, or from the display screen using a cursor. This is time-consuming (often hours) and error-prone, and limited by the precision of the display or ruler. What is often not realised is that the data themselves are held to much higher precision (usually 0.0-0.01 pixels) in the PDF document. PDF display tools are required to read, transform and display vector primitives ("Paths") and by using suitable software we can extract the coordinates without ambiguity or loss of inherent precision. A plot of, say, 10000 x-values onto 500 pixels will suffer massive overlap in the display but the data are recoverable from the PDF.

@@

Based on the plot body, absolute locations of the individual data points are extracted. Not all vector images are created in a similar way, but in the simplest scenario with data points depicted as circles, the vector gives three parameters: the $x$ coordinate of the centre, the $y$ coordinate of the centre, and the radius $r$. As such, for a simple circle the underlying vector code (in Scalable Vector Graphics, SVG)  might look as follows:
```html
<circle cx="103.71" cy="121.22" r="25.234" fill-opacity="0" stroke="#cf1d35" stroke-width=".26458"/>
```
This information can be readily extracted after isolating the vector figure from a PDF file. The current alpha software is primarily developed to operate on circles of similar size within one plot but can be extended for data depicted in other ways.

PMR  'norma` interprets characters as "ladders" of numeric values along the axes, and probable axial titles. Where necessary non-horizontal text is rotated. It then identifies a rectangular box, examines it for tick marks and matches the ticks to the axial scale values. This leads to high-precision scale and offset values for transforming screen to user coordinates.

In order to make the absolute locations of the shapes represent the original data points as accurately as possible, they are mapped onto the identified x- and y-axis. Although absolute locations retain the relative relations between the individual data points, they are not representative of the original data.  Subsequently, the location of the x-axis and y-axis, combined with the information about the scale, are combined to remap the absolute locations on the canvas into the original data points. The current alpha software assumes a linear scale, but logarithmic scales could be incorporated at a future stage.
<!-- verify with PMR AGREED -->

@@

In order to extract data from vector figures with the software `norma`, `r stepsnr` steps are taken. First, the user needs to organize all original PDFs into one folder. Second, this folder needs to be converted to a `cproject` structure. The `cproject` structure normalizes the contents for each paper into a `ctree`, such that subsequent operations are trivial to standardize (and extensions can be applied relatively easily). For example, the root folder might contain `ctree1.pdf`, but after transforming the root folder into a `cproject` it contains a folder `ctree1/` with `fulltext.pdf`. By running the command 
```bash
java -jar bin/norma-0.5.0-SNAPSHOT-jar-with-dependencies.jar --project corpus-raw  --fileFilter '.*/(.*).pdf' --makeProject '(\1)/fulltext.pdf'
```
the folder `corpus-raw` (`--project corpus-raw`) is restructured into a `cproject` structure, containing a folder for each PDF file (`--fileFilter '.*/(.*).pdf' --makeProject '(\1)/fulltext.pdf'`). This results in the following folder structure, where `fulltext.pdf` is the original PDF and `ctree1` (etc.) may captute the DOI or other document identifiers.:
```bash
cproject/
├── ctree1
│   └── fulltext.pdf
├── ctree2
│   └── fulltext.pdf
...
└─── ctreeN
    └── fulltext.pdf
```

@@

As the results indicate, vector figures are a fruitful and feasible resource for data extraction. Based on initial alpha software of `norma` to extract data from vector figures, we correctly extracted data from `r round((dim(vec_dat[vec_dat$x_ax_mapped == 'yes' & vec_dat$y_ax_mapped == 'yes' & vec_dat$datapoints_extracted == vec_dat$datapoints_manual, ])[1] / dim(vec_dat)[1]) * 100, 0)`% of funnel plots for which data was extracted. This is a very strict assessment, considering our manual investigation depicted that four extracted datasets (which used a non-standard direction of the y-axis) only required a simple reversal of one axis to be 100% accurate (i.e., 1.23 should be -1.23, etc.), adding a constant to all data points on an axis to adjust an incorrect mapping, or by rescaling an axis to fit to the logarithmic scale of one axis. If these manual corrections for four funnel plots are made, `r round(((dim(vec_dat[vec_dat$x_ax_mapped == 'yes' & vec_dat$y_ax_mapped == 'yes' & vec_dat$datapoints_extracted == vec_dat$datapoints_manual, ])[1] + 4) / dim(vec_dat)[1]) * 100, 0)`% of all funnel plots for which data were correctly extracted.
